What Drives Emergency Medical Services (EMS) Demand?

### 2024-09-27 paper/paper.qmd
Edited the entire paper and rendered a new pdf version.

---
title: "What Drives Emergency Medical Services (EMS) Demand?"
subtitle: "Community-based EMS-demand-management strategies that focus on medical emergencies, emergency transfers, and vehicle accidents may help enhance paramedic services"
author: 
  - Julia Lee
thanks: "Code and data are available at: https://github.com/JuliaJLee/Toronto_Paramedic_Services.git"
date: today
date-format: long
abstract: "By examining Paramedic Services Incident Data (2017-2022) provided by Toronto Paramedic Services, the following anaylsis aims to identify the key drivers of Emergency Medical Services (EMS) demand to help inform demand-management strategies. Through data visualization, this analysis finds that medical emergencies, incidents requiring emergency transfers, and vehicle accidents appear to be the most prominent drivers of EMS demand. This suggests that strategies that target these drivers may enhance the overall efficiency of paramedic services. Additionally, the present analysis emphasizes that thoughtful consideration for the unique needs of those who require EMS is important to build practical and productive EMS-demand-management strategies."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

"In case of an emergency, please call 911." From building walls to road signs, this is a message that appears almost everywhere, and it instills the idea that no matter the circumstance, there will always be emergency services available to help. However, research shows that this is not the case. A report on Toronto's paramedic services highlights that there were no ambulances available to over 1,000 calls in 2023 [@auditorreport]. With drastically increasing wait times for those who required Emergency Medical Services (EMS), it is crucial to understand why paramedic services are unable to keep up with a seemingly growing demand for EMS.

In response to this rising challenge, the report posits that in order to elevate the "efficiency and effectiveness" [@auditorreport] of Toronto's paramedic services, further research on how to manage EMS demand is needed. Thus, this paper seeks to investigate the factors that seem to be driving EMS demand. Using Paramedic Services Incident Data offered by Toronto Paramedic Services, the aim is to identify which types of emergencies (e.g. medical emergencies or vehicle accidents) drive EMS demand and examine whether . Knowing this can help inform effective strategies that allocate resources to these drivers for better EMS demand management.

With this, this analysis first examines the total number of emergency calls made from 2017 to 2022 to better understand how EMS demand has changed over time. Then, it looks at EMS demand through five different incident types as outlined by Toronto Paramedic Services to identify the incident types that push EMS demand. Lastly, this analysis considers whether drivers of EMS demand differ between different regions. As a result, the present analysis finds that overall, medical emergencies, incidents requiring emergency transfers, and vehicle accidents seem to be the most prominent drivers of EMS demand. However, as different drivers of EMS demand are found for different regions, this analysis highlights that while strategies that target medical emergencies, incidents requiring emergency transfers, and vehicle accidents may enhance the overall efficiency of paramedic services, it is also crucial to consider the unique needs of various regions and their residents to shape stronger EMS-demand-management strategies.

In what follows, a description of the data that was used within this analysis (@sec-data), detailed account of the results (@sec-results), and a brief discussion of the analysis along with its limitations and next steps (@sec-discussion) will be provided.

@@

To further understand what drives EMS demand, this analysis uses the Forward Sortation Areas (FSA) provided by the Paramedic Services Incident Data (2017-2022) to map EMS demand across all the locations Toronto Paramedic Services respond to. The map is based on the Paramedic Services Incident Data from 2022. By grouping the data shown in @sec-analysis by FSAs, the types of incidents recorded within each unique FSA and the number of incidents that occurred with each incident type could be found. The "driver" of EMS demand for a given FSA was determined by the incident type that had the most number of calls. A small snippet of the modified analysis data is shown below in @tbl-mapdata. The column, "CFSAUID" contains every FSA that Toronto Paramedic Services responded to in 2022.

@@

To map this data, a shapefile with the spatial information for all recorded FSAs is combined with the data shown in @tbl-mapdata. The shapefile, titled, "2021 Census - Boundary Files", is provided by the Open Government Data portal from the Government of Canada [@2021census]. Additional information about this dataset, including how it is used can be found in the Appendix (@sec-appendix).

@@

## EMS Demand from 2017 to 2022 Based on Incident Type

To further unpack what is driving this consistent demand, the calls for individual incident type are also considered. @fig-incidentemsdemand illustrates the number of emergency calls for each incident type along with the total number of emergency calls. Based on @fig-incidentemsdemand, demands for EMS are markedly higher for medical emergencies than any other incident type, suggesting that medical circumstances are key drivers that propel EMS demand.

@@

Similarly, as vehicle accidents and emergency transfers can be potential drivers of EMS demand, overall EMS demand may be better managed if paramedic service strategies also target vehicle-accident- and emergency-transfer-related calls. Traffic congestion has been found to increase paramedic response times [@journal], and with the current increase in traffic due to multiple construction projects throughout Toronto, longer paramedic response times and the unavailability of paramedic services can be expected. To efficiently navigate EMS demand for vehicle accidents and emergency transfers, strategies may center around vehicle accident prevention or the allocation of more staff and resources to accident or emergency-prone regions for faster response times.

@@

By examining EMS demand through different incident types, the investigation above finds that medical emergencies are responsible for a significantly large portion of 911 calls. However, this finding is a generalization of all the forward sortation areas that Toronto Paramedic Services responded to over several years. Medical emergencies may not be the driver of EMS demand for all regions or communities. Thus, the following part of this analysis looks to identify the driver for every unique forward sortation area Toronto Paramedic Services responded to in 2022. If different forward sortation areas have different drivers of EMS demand, then strategies that are tailored to each region's circumstance may be more effective in managing overall EMS demand.

@@

```{r, results = 'hide'}
#| message: false
#| echo: false 
#| warning: false
#| label: fig-map
#| fig-cap: EMS Demand Across Forward Sortation Areas (Based on 2022 data)

@@

forward_sortation_areas <- st_read("~/Toronto_Paramedic_Services/data/ForwardSortationAreas/Forward_SortationAreas")
view(forward_sortation_areas)

@@

By focusing in on the city of Toronto with @fig-mapzoom, EMS demand seems to be significantly driven by medical emergencies and slightly driven by vehicle accidents and emergency transfers.

These results suggest that what drives EMS demand is dependent on the characteristics of a particular region. Socio-economic or demographic factors such as income or population density of a given area can impact the types of incidents that occur, total number of emergency calls that are made and the severity of each incident. Thus, there is a need to consider individual communities to build tailored solutions that effectively address their unique needs and better manage widespread EMS demand.

@@

As high demand for EMS continues to be a challenge, identifying the areas where most of the demand comes from can help to inform effective demand-management strategies. Through an investigation of EMS demand, the present analysis identifies three main drivers of EMS demand (medical emergencies, incidents requiring emergency transfers, and vehicle accidents) and also considers how these drivers might differ in different contexts (i.e. different regions or communities).

To further this research, the relationship between the driver of EMS demand for every forward sortation area and each area's socio-economic or demographic characteristics should be examined. As there are many variables that can impact one's health and ultimately, whether or not they have to seek EMS, understanding how factors like income or population density affect the occurrence of medical emergencies, emergency transfers, and vehicle accidents can help build tailored and effective strategies. It may also be helpful to consider ambulance station locations in addition to @fig-map to see whether ambulance stations and resources are optimally distributed and if additional resources are needed depending on the EMS demand driver for a particular region.

@@

Though this analysis carefully examines EMS demand, there are some limitations to consider. It is mentioned in @sec-analysis that the records of incidents without an assigned incident type were omitted from the analysis. However, this may have limited the overall representation of each incident type as well as the conclusions that are made. Further, it is important to note that @fig-map only data from 2022. So, although @fig-map demonstrates that differences between regions and their drivers for EMS demand can exist, it is not an accurate reflection of what each forward sortation area because only one instance (the year 2022) is considered.

@@

summary_data |>
    kable(
    col.names = c("Medical Emergencies", "Vehicle Accidents", "Emergency Transfers",
                  "Fires", "Airport Standbys", "Unclassified", "Total Calls"),
    digits = 1,
    booktabs = TRUE,
    linesep = "",
    align = c("l", "l", "l", "l", "l", "l"),
    format.args = list(big.mark = ",")
  )

@@

data |>
  slice(1:3) |>
  kable(
    col.names = c("CFSAUID", "DGUID", "PRUID", "PRNAME", 
                  "LANDAREA", "Geometry"),
    digits = 1,
    booktabs = TRUE,
    linesep = "",
    align = c("l", "l", "l", "l", "l", "l"),
    format.args = list(big.mark = ",")
  )
```
The shapefile that is used in this analysis contains the forward sortation area boundaries for every province and territory of Canada. In this file, every forward sortation area is matched with a set of spatial information (i.e. coordinates). @tbl-shape illustrates this data. "CFSAUID" contains every FSA that Toronto Paramedic Services responded to in 2022, and "DGUID" refers to the "Dissemination Geography Unique Identifier (DGUID)" - an ID that all geographic areas in Canada have. The variables, "PRUID" and "PRNAME", represent province numbers and names, respectively. "LANDAREA" is the total area in square kilometers of each forward sortation area, and "Geometry" holds the spatial information for every area.

>>>>

---
title: "What Drives Emergency Medical Services (EMS) Demand?"
subtitle: "Community-based paramedic service strategies that focus on medical emergencies, emergency transfers, and vehicle accidents may help to effectively manage EMS demand"
author: 
  - Julia Lee
thanks: "Code and data are available at: https://github.com/JuliaJLee/Toronto_Paramedic_Services.git"
date: today
date-format: long
abstract: "By examining Paramedic Services Incident Data (2017-2022) provided by Toronto Paramedic Services, the following anaylsis aims to identify the key drivers of Emergency Medical Services (EMS) demand to help inform demand-management strategies. Through data visualization, this analysis finds that medical emergencies, incidents requiring emergency transfers, and vehicle accidents are overall drivers of EMS demand. This suggests that centering strategies around these drivers may enhance the overall efficiency of paramedic services. Additionally, this analysis emphasizes that thoughtful consideration for the unique needs of those who require EMS is important to build practical EMS-demand-management strategies."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

"In case of an emergency, please call 911." From building walls to road signs, this is a message that appears almost everywhere, and it instills the idea that no matter the circumstance, there will always be emergency services available to help. However, research shows that this is not the case. A report on Toronto's paramedic services highlights that there were no ambulances available for over 1,000 calls in 2023 [@auditorreport]. With drastically increasing wait times for those who required Emergency Medical Services (EMS), it is necessary to understand why paramedic services are unable to keep up with a seemingly growing demand for EMS.

In response to this rising challenge, the report posits that in order to elevate the "efficiency and effectiveness" [@auditorreport] of Toronto's paramedic services, further research on how to manage EMS demand is needed. Thus, by using Paramedic Services Incident Data offered by Toronto Paramedic Services, this analysis seeks to identify which types of emergencies (e.g. medical emergencies or vehicle accidents) drive EMS demand. Knowing the drivers of EMS demand can help inform paramedic service strategies that allow for better EMS demand management.

With this, this analysis first examines the total number of emergency calls made from 2017 to 2022 to better understand how EMS demand has changed over time. Then, it looks at EMS demand through five different incident types as outlined by Toronto Paramedic Services to identify the drivers that push EMS demand. Lastly, this analysis considers whether drivers of EMS demand differ across different communities. As a result, the present analysis finds that overall, medical emergencies, incidents requiring emergency transfers, and vehicle accidents are the key drivers of EMS demand. However, not all communities are found to have the same drivers. So, while strategies that target medical emergencies, incidents requiring emergency transfers, and vehicle accidents may enhance the overall efficiency of paramedic services, it is important to consider the unique needs of various communities to shape stronger EMS-demand-management strategies.

In what follows, a description of the data that was used within this analysis (@sec-data), a detailed account of the results (@sec-results), and a brief discussion of the analysis (@sec-discussion) is provided.

@@

To further understand what drives EMS demand, this analysis uses the Forward Sortation Areas (FSA) provided by the Paramedic Services Incident Data (2017-2022) to map EMS demand across all the locations Toronto Paramedic Services respond to. The map is based on the Paramedic Services Incident Data from 2022. By grouping the data shown in @sec-analysis by FSAs, the types of incidents recorded within each unique FSA and the number of incidents that occurred with each incident type could be found. The "driver" of EMS demand for a given FSA was determined by the incident type that had the most number of calls. This data is shown below in @tbl-mapdata. The column, "CFSAUID" contains every FSA that Toronto Paramedic Services responded to in 2022.

@@

To map this data, a shapefile with the spatial information for all recorded FSAs is combined with the data shown in @tbl-mapdata. The shapefile from the 2021 Census is provided by the Open Government Data portal from the Government of Canada [@2021census]. Additional information about this dataset, including how it is used can be found in the Appendix (@sec-appendix).

@@

## EMS Demand from 2017 to 2022 Based on Incident Type {#sec-incidenttypes}

To further unpack what is driving this consistent demand, the number of emergency calls for individual incident type are also considered. @fig-incidentemsdemand illustrates the number of emergency calls for each incident type along with the total number of emergency calls. Based on @fig-incidentemsdemand, demands for EMS are markedly higher for medical emergencies than any other incident type, suggesting that medical circumstances are key drivers that propel EMS demand.

@@

Similarly, as vehicle accidents and emergency transfers can be potential drivers of EMS demand, overall EMS demand may be better managed if paramedic service strategies also target vehicle accident and emergency transfer-related calls. Traffic congestion has been found to increase paramedic response times [@journal], and with the current increase in traffic due to multiple construction projects throughout Toronto, longer paramedic response times and the unavailability of paramedic services can be expected. To efficiently navigate EMS demand for vehicle accidents and emergency transfers, strategies may center around vehicle accident prevention or the allocation of more staff and resources to accident or emergency-prone regions for faster response times.

@@

By examining EMS demand through different incident types, the investigation above finds that medical emergencies are responsible for a significantly large portion of 911 calls. However, this finding is a generalization of all the forward sortation areas that Toronto Paramedic Services responded to over several years. Medical emergencies may not be the main driver of EMS demand for all communities. Thus, the following part of this analysis looks to identify the driver for every unique forward sortation area Toronto Paramedic Services have responded to in 2022. If different forward sortation areas have different drivers of EMS demand, then strategies that are tailored to each community's circumstance may be more effective in managing overall EMS demand.

@@

```{r, results = 'hide'}
#| message: false
#| echo: false 
#| warning: false
#| label: fig-map
#| fig-cap: EMS Demand Across Forward Sortation Areas (Based on 2022 Data)

@@

@@

By focusing in on the city of Toronto with @fig-mapzoom, EMS demand in many communities seems to be significantly driven by medical emergencies, and there are only a few communities whose EMS demand is driven by vehicle accidents or emergency transfers.

These results support the findings from @sec-incidenttypes. Medical emergencies, emergency transfers, and vehicle accidents are shown to be relevant drivers of EMS demand. The findings further imply that what drives EMS demand is dependent on the characteristics of a particular region. Socio-economic or demographic factors such as income or population density of a given area can impact the types of incidents that occur, total number of emergency calls that are made and the severity of each incident. Given this, general strategies that target the three main drivers of EMS demand may not be effective for all communities. Thus, there is a need to consider the characteristics of different communities to build community-based strategies that effectively address the main drivers of EMS demand.

@@

As high demand for EMS continues to be a challenge, identifying the areas where most of the demand comes from can help to inform effective demand-management strategies. Through an investigation of EMS demand, the present analysis identifies three main drivers of EMS demand (medical emergencies, incidents requiring emergency transfers, and vehicle accidents) and also considers how these drivers might differ across various communities.

To further this research, the relationship between the driver of EMS demand and socio-economic or demographic characteristics for each forward sortation area should be examined. As there are many variables that can impact one's health and ultimately, whether or not they have to seek EMS, understanding how factors like income or population density affect the occurrence of medical emergencies, emergency transfers, and vehicle accidents can allow for the formation of effective community-based strategies. It may also be helpful to consider ambulance station locations in addition to @fig-map to see whether ambulance stations and resources are optimally distributed and if additional resources are needed depending on the EMS demand driver for a particular community.

@@

Though this analysis carefully examines EMS demand, there are some limitations to consider. It is mentioned in @sec-analysis that the records of incidents without an assigned incident type were omitted from the analysis. This may have limited the overall representation of the number of emergency calls for each incident type. Further, it is important to note that @fig-map only considers data from 2022. So, although @fig-map demonstrates that differences between communities and their drivers for EMS demand can exist, a better reflection of the actual drivers of EMS demand for each community may be achieved by considering multiple years of incident data (i.e. from 2017 to 2022).

@@

summary_data |>
    kable(
    col.names = c("Medical Emergencies", "Vehicle Accidents", "Emergency Transfers",
                  "Fires", "Airport Standbys", "Unclassified", "Total Calls"),
    digits = 1,
    booktabs = TRUE,
    linesep = "",
    align = c("l", "l", "l", "l", "l", "c"),
    format.args = list(big.mark = ",")
  )

@@

data |>
  slice(1:3) |>
  kable(
    col.names = c("CFSAUID", "DGUID", "PRUID", "PRNAME", 
                  "LANDAREA", "Geometry"),
    digits = 1,
    booktabs = TRUE,
    linesep = "",
    align = c("l", "l", "l", "l", "l", "c"),
    format.args = list(big.mark = ",")
  )
```
The shapefile that is used in this analysis contains the forward sortation area boundaries for every province and territory of Canada. In this file, every forward sortation area is matched with spatial information (i.e. coordinates). @tbl-shape illustrates this data. "CFSAUID" contains every FSA that Toronto Paramedic Services responded to in 2022, and "DGUID" refers to the "Dissemination Geography Unique Identifier (DGUID)" - an ID that all geographic areas in Canada have. The variables, "PRUID" and "PRNAME", represent province numbers and names, respectively. "LANDAREA" is the total area in square kilometers of each forward sortation area, and "Geometry" holds the spatial information for every area.
